---
title: "General Linear Hypothesis Tests For Quantile Regression Models Using qrinf"
author: "Nathan Laroy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
header-includes:
   - \usepackage{bm}
vignette: >
  %\VignetteIndexEntry{General Linear Hypothesis Tests For Quantile Regression Models Using qrinf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## General Linear Hypotheses

The quantile regression framework, like its linear regression counterpart, allows to test general linear hypotheses of various sorts using Wald-type statistics. Let a multiple quantile regression model take the following form:

$$
Q_Y(\tau\mid \boldsymbol{X}) = \boldsymbol{X\beta}(\tau),
$$

with $\mathbf{X}$ a covariate row vector of length $p+1$, and $\boldsymbol{\beta}(\tau)$ a coefficient column vector of length $p+1$, first element $\beta_0(\tau)$. Consequently, define

$$
\boldsymbol{\xi} = \left(\boldsymbol{\beta}(\tau_1)', \dots, \boldsymbol{\beta}(\tau_m)'\right)',
$$

and $\boldsymbol{\beta}(\tau_j)=\left(\beta_0(\tau_j), \dots, \beta_p(\tau_j)\right)$. One may formulate two types of general linear hypotheses based on this model. The first concerns the extended coefficient vector $\boldsymbol{\xi}$ and takes the canonical form:

$$
H_0: \boldsymbol{R\xi}=\boldsymbol{r}.
$$

General linear hypotheses of this form allow to assess verbal hypotheses such as "the first and third quartile regression of $Y$ on $\mathbf{X}$ yield identical coefficients", or "the population value of $\beta_2(\tau)$ for the 90th percentile regression is exactly double that of the 10th percentile regression", and any combination of these statements. They involve *point estimates* of quantile regression coefficients for a finite set of quantile criteria (i.e., they involve linear combinations of $\beta_k(\tau=\tau_{0,l})$, for $k\in\{0,\dots,p\}$ and $\tau_{0,l}$ selected quantile criteria under the null hypothesis). The appropriate Wald-type statistic takes the following form:

$$
T = \left(\boldsymbol{R}\hat{\boldsymbol{\xi}}-\boldsymbol{r}\right)'\left[\boldsymbol{R}\hat{\boldsymbol{V}}\boldsymbol{R}'\right]^{-1}\left(\boldsymbol{R}\hat{\boldsymbol{\xi}}-\boldsymbol{r}\right),
$$

which converges asymptotically to a $\chi^2_q$ distribution with $q$ the rank of $\boldsymbol{R}$. A practically challenging aspect of conducting a Wald-type inferential test for this type of null hypothesis pertains to the estimation of the asymptotic covariance of $\boldsymbol{R}\hat{\boldsymbol\xi}-\boldsymbol{r}$, i.e., $\boldsymbol{R}\hat{\boldsymbol{V}}\boldsymbol{R}'$. Commonly tested null hypotheses, such as, e.g., $H_0: \beta_k(\tau)=0$ and $H_0: \boldsymbol{\beta}(\tau)=\boldsymbol{0}$, have already been implemented in `quantreg` through existing convenience functions (e.g., `anova`). The `qrinf` package was specifically developed to simplify matters for applied researchers whenever they wish to assess a null hypothesis which does not enjoy a standard implementation in `quantreg`, and so for which the asymptotic covariance matrix would have to be manually estimated and the statistic manually obtained.

A second type of general linear hypothesis takes the following form:

$$
H_0: \boldsymbol{R\beta}(\tau)=\boldsymbol{r}(\tau),
$$

with $\boldsymbol{r}(\tau)$ a vector-valued function of $\tau$. Contrary to the previous case, general linear hypotheses pertain here to continuous closed intervals of quantiles (i.e., $\tau\in[\tau_a,\tau_b]\subsetneq(0,1)$). They allow to assess verbal hypotheses which pertain to the quantile regression process, e.g., "$\beta_3(\tau)$ is constant in the interval $[0.5, 0.95]$", or "$\beta_1(\tau)$ changes sign, but not magnitude, for $\tau>0.5$ compared to $\tau<0.5$". The appropriate sup-Wald statistic takes the following form:

$$
T = \sup \left(\boldsymbol{R}\hat{\boldsymbol\beta}(\tau)-\boldsymbol{r}(\tau)\right)'\left[\boldsymbol{R}\hat{\boldsymbol{V}}(\tau,\tau)\boldsymbol{R}'\right]^{-1}\left(\boldsymbol{R}\hat{\boldsymbol\beta}(\tau)-\boldsymbol{r}(\tau)\right),
$$

which behaves asymptotically like the $\sup$ of a tied-down Bessel process of order $q$, with $q$ the rank of $\boldsymbol{R}$. As before, `quantreg` includes some convenient implementations of specific instances of general linear hypotheses concerning the quantile regression process (namely, those which assess null hypotheses of location-shift and location-scale, via the `khmaladzeTest` function). However, researchers may be interested in testing a verbal hypothesis which cannot be rephrased as one which concerns location-shift or location-scale models, in which case the sup-Wald statistic would have to be manually calculated. The `qrinf` package allows applied researchers to conduct a hypothesis of the above sort, without having to write a bespoke script.

The remainder of this vignette describes `glhtxi` and `glhtprocess`, the work horses of each hypothesis testing procedure, in more detail, providing code examples of their usage.

## `glhtxi`: general linear hypothesis tests on $\xi$.

### example 1.1

In order to make use of the `glhtxi` function, one first loads the `quantreg` and `qrinf` packages (after having installed them). The former can be installed using `install.packages("quantreg")`, while the latter is best installed using `pak::pak("nathanlaroy/qrinf")`.

```{r}
# Load
library(qrinf)
```

We will rely on the `mtcars` data set to illustrate the uses of `glhtxi`. Below is part of the `data.frame`.

```{r}
knitr::kable(head(mtcars))
```

Suppose we wish to model the miles/(US) gallon (`mpg`) of the cars in function of number of cylinders (`cyl`), displacement (cu.in.) (`disp`) and gross horsepower (`hp`). If we estimate a quantile regression model for a dense grid of quantile criteria and plot the resulting sample paths of the coefficient estimators, we find the following:

```{r, fig.width=5, fig.height=5}
rqmod <- rq(mpg ~ cyl + disp + hp, tau = seq(0.05, 0.95, 0.01), data = mtcars)
plot(rqmod)
```

Suppose the literature states that the effect of `cyl` is twice in magnitude for the 80th percentile as compared to the 20th percentile, controlled for `disp` and `hp`. Based on the visualization, it seems like this could be the case, but we are skeptical and wish to quantify the statistical evidence against this statement. To that end, we can rephrase the verbal hypothesis as a linear restriction on $\boldsymbol{\xi}$. Specifically, the null hypothesis becomes:

$$
\begin{align*}
H_0: &&2\beta_{\text{cyl}}(0.2) &= \beta_{\text{cyl}}(0.8)\\
\iff&&2\beta_1(0.2) - \beta_1(0.8) &= 0
\end{align*}
$$

In order to proceed, we need to translate the above symbolic representation into a restriction matrix $\boldsymbol{R}$, so that $\boldsymbol{R\xi}$ represents the left-hand side of the null equation, and a null identity matrix $\boldsymbol{r}$ representing its right-hand side. From the model, we know that $\boldsymbol{\xi}=(\beta_0(0.2), \beta_1(0.2),\dots,\beta_2(0.8),\beta_3(0.8))'$, 8 elements long (4 coefficients times 2 quantile criteria), so $\boldsymbol{R}$ becomes:

```{r}
R <- rbind(c(0, 2, 0, 0,
             0, -1, 0, 0))
r <- matrix(c(0))
```

The `glhtxi` function requires further that we separately specify the $\tau$-quantiles of interest in a simple vector, and a separate `formula` object of the model. We can extract the latter from the previously run instance of `rq`.

```{r}
taus <- c(0.2, 0.8)
form <- formula(rqmod)
```

We may now run `glhtxi`, which will efficiently assemble the respective asymptotic covariance estimate and return a list of values. We leave the argument `se` unspecified, so it defaults to `"nid"` (i.e., errors assumed non-identically, independently distributed).

```{r}
glhtxi(taus = taus, form = form, data = mtcars,
       R = R, r = r, se = "nid")
```

In the output, we are given a copy of the restriction matrix `$R`, the realized identity under the null `$H0` (i.e., $\boldsymbol{R}\hat{\boldsymbol{\xi}}-\boldsymbol{r}$), the observed test statistic `Tn`, the degrees of freedom `q`, and the p-value `p.value`. Here, $p=0.119$, so we fail to reject the null hypothesis.

We also see that the function has returned a series of warning messages. These all derive from `summary.rq`, which is called under the hood iteratively to calculate blocks of the covariance matrix. They indicate that a non-positive density was estimated, which the function dealt with under the hood (i.e., densities cannot be negative). Too many such warnings can indicate a model misspecification problem, but a small amount is generally not problematic.

### example 1.2

We next illustrate a second example where the null hypothesis consists of several identities. For example, assume the literature states that the effect of each independent variable is exactly 0.25 units smaller for the 25th versus the 75th conditional quantile, conditional on the others. That is, the following set of null hypothesis identities holds:

$$
H_0:
\begin{cases}
  \beta_1(0.25) - \beta_1(0.75) = -0.25\\
  \beta_2(0.25) - \beta_2(0.75) = -0.25\\
  \beta_3(0.25) - \beta_3(0.75) = -0.25
\end{cases}
$$

We can translate this into the respective `R` and `r` matrices as follows, setting `taus` as well, and retaking `form` from before:

```{r}
R <- rbind(c(0, 1, 0, 0, 0, -1, 0, 0),
           c(0, 0, 1, 0, 0, 0, -1, 0),
           c(0, 0, 0, 1, 0, 0, 0, -1))
r <- cbind(c(-0.25, -0.25, -0.25))
taus <- c(0.25, 0.75)

glhtxi(taus = taus, form = form, data = mtcars, R = R, r = r)
```

We find very strong evidence against this null hypothesis, and so choose to reject it. Based on the observed null identity `$H0`, we see that, indeed, none of the equalities are particularly close to 0, which should be the case if the null hypothesis actually holds true. The p-value is practically nil.

## `ghltprocess`: general linear hypothesis tests on $\boldsymbol{\beta}(\tau)$

In this section, use of `glhtprocess` is illustrated.

### example 2.1

Based on the plot from before, one might wonder if the effect of `hp` is nil for all quantiles in the unit interval, conditional on `cyl` and `disp`. To test this, we use `glhtprocess`. As with `glhtxi`, the function requires that one specifies the restriction matrix $\boldsymbol{R}$ and null identity $\boldsymbol{r}$ (which here takes the form of a function). $\boldsymbol{R}$ and $\boldsymbol{r}$ have dimensions such that $\boldsymbol{R\beta}(\tau)-\boldsymbol{r}$ is a legal matrix operation (i.e., $\boldsymbol{R}$ is a $q\times(p+1)$ matrix and $\boldsymbol{r}$ is a $q\times1$ matrix). We entertain the verbal hypothesis which states that $\beta_{\text{hp}}(\tau)$ is constant in $\tau$, or symbolically, $\beta_{\text{hp}}(\tau)=0$. We also need to specify the bounds of the continuous closed quantile interval of interest. We take $[0.05, 0.95]$, trimming off the tails to account for the high uncertainty.

```{r, eval=FALSE}
R <- rbind(c(0, 0, 0, 1))
r <- matrix(c(0))
taus <- c(0.05, 0.95)

glhtprocess(taus = taus, form = form, data = mtcars, R = R, r = r)
```
```{r, echo=FALSE}
R <- rbind(c(0, 0, 0, 1))
r <- matrix(c(0))
taus <- c(0.05, 0.95)

suppressWarnings(processtest <- glhtprocess(taus = taus, form = form, data = mtcars, R = R, r = r))
processtest
```

We find that the obtained statistic is `r round(processtest$Tn,2)`, which is smaller than the critical value associated with the 10% significance level. So, we fail to reject the null hypothesis at said significance level. From the output, we find that `$interpolated` is `FALSE`, meaning that the obtained value for `$lambda` appears in the tables by Estrella (2003) and so the critical values are exact. When `$interpolated` returns `TRUE`, it means that `$lambda` does not appear in the tables by Estrella (2003). Consequently, the critical values for the different significance levels are linearly interpolated, yielding approximate values.

Note that a series of warning messages was omitted from this vignette for the purpose of readability. Users should be aware that `glhtprocess` is prone to returning many warnings due to the consecutive estimation of multiple asymptotic covariance blocks under the hood using `quantreg::summary.rq`. They can be suppressed by wrapping the call inside `suppressWarnings()`, but this is generally not advised. A great number of warnings can point to model specification issues, and so should be investigated rather than ignored.

### example 2.2

As with `glhtxi`, the function `glhtprocess` allows to test combined null hypotheses. We illustrate this for the verbal hypothesis that the effects of `disp` and `cyl` are both different from zero for the interval $\tau\in[0.05, 0.5]$, controlled for `cyl`. Symbolically, this may be written as follows:

$$
H_0:
\begin{cases}
\beta_1(\tau) = 0\\
\beta_2(\tau) = 0
\end{cases}
$$

We define $\boldsymbol{R}$ and $\boldsymbol{r}(\tau)$, and set `taus` to the desired interval, in the code chunk below, and run `ghltprocess`.

```{r, eval=FALSE}
R <- rbind(c(0, 0, 1, 0),
           c(0, 0, 0, 1))
r <- matrix(c(0, 0))
taus <- c(0.05, 0.5)

glhtprocess(taus = taus, form = form, data = mtcars, R = R, r = r)
```
```{r, echo=FALSE}
R <- rbind(c(0, 0, 1, 0),
           c(0, 0, 0, 1))
r <- matrix(c(0, 0))
taus <- c(0.05, 0.5)

suppressWarnings(processtest <- glhtprocess(taus = taus, form = form, data = mtcars, R = R, r = r))
processtest
```

We find that we can convincingly reject the null hypothesis. The obtained test statistic is `r round(processtest$Tn, 2)`, which is far greater than the critical value associated with the 1% significance level. Based on the output, we find that `$lambda` is `r processtest$lambda`, which does not appear in the Estrella tables. Hence, the critical values were calculated by means of linear interpolation (`$interpolated` is `TRUE`). However, because the obtained test statistic is nowhere near the interpolated values, we can rest assured that the rejection of the null hypothesis is legitimate.

Note that, again, we suppressed warnings for the purpose of the readability of this vignette and so the focus could be on the output itself. In making use of `glhtprocess`, one should not blatantly ignore such warnings.

# References

Estrella, A. (2003). Critical values and p values of bessel process distributions: Computation and
application to structural break tests. *Econometric Theory, 19*(6), 1128–1143.
https://doi.org/10.1017/S0266466603196107